Hierarchical model assessing the impacts of time to detection distribution assumptions on detection probability estimation

### 2016-09-08 Draft.tex
Make minor changes
including ref to Borchers preprint

\documentclass[useAMS,usenatbib,referee,12pt]{article}
\usepackage[margin=1.0in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{parskip}
\usepackage[round]{natbib}
\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage{caption}
\usepackage{tabularx, booktabs}
\usepackage{float}
\usepackage{adjustbox}
\usepackage[toc,page]{appendix}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{xr}
\externaldocument[S-]{Supplement}
\doublespacing

@@

Abundance estimates from animal point-count surveys require accurate estimates of detection probabilities.  
The standard model for estimating detection from removal-sampled point-count surveys assumes that organisms at a survey site are detected at a constant rate; however, this assumption is often not justified.  
%Detection rates can be influenced by organismal behaviors, survey methodology, and observer effort.  
%Failure to account for non-constant detection leads to biased estimates of detection and therefore of abundance.  
We consider a class of N-mixture model that allows for detection heterogeneity over time through a flexibly defined time-to-detection distribution (TTDD) and allows for fixed and random effects for both abundance and detection.
%We consider a class of N-mixture models that allow for detection heterogeniety through a mixture component, flexible two-parameter time-to-detection distributions (TTDDs), and allows for fixed and random effects for both abundance and detection.
%Whereas, under the standard approach, detection probability is modeled by dividing the observation period into equal-duration intervals, we instead model the detection {\it rate} in continuous time via a time-to-detection distribution (TTDD) embedded within a hierarchical N-mixture framework.  
Our model is thus a combination of survival time-to-event analysis with unknown-N, unknown-p abundance estimation.  
We specifically explore two-parameter families of TTDDs (e.g. gamma) that can additionally include a mixture component to model increased probability of detection in the initial observation period.
We find that modeling a TTDD by using a mixture component is necessary when data have a chance of arising from a distribution of this nature.
In addition, two-parameter distributions can outperform exponential-based models both when the truth is exponential or non-exponential.
Finally, we analyze an Overbird data set from the Chippewa National Forest using mixed effect models for both abundance and detection.
We demonstrate that the effects of explanatory variables on abundance and detection are consistent across mixture TTDDs but that flexible TTDDs result in lower estimated probabilities of detection and therefore higher estimates of abundance. 
%We find that modeling a TTDD by using a mixture component for increased probability of detection in the initial observation period is necessary when data have a chance of arising from distributions of this nature.
%In addition, the flexibility offered by two-parameter distributions, e.g. gamma, can outperform exponential based models when the truth is exponential or non-exponential.
%Finally, we analyze an Overbird data set from the Chippewa National Forest using mixed effect models for both abundance and detection and demonstrate that the effects of explanatory variables on abundance and detection are consistent across mixture TTDDs and that flexible TTDDs result in lower estimated probabilities of detection and therefore higher estimates of abundance. 
% We apply this model to Ovenbird counts from the Chippewa National Forest and to datasets simulated under different time-to-detection patterns.  
% Models assuming constant detection rates produce biased estimates of detection when true detection rates vary with time, whereas models allowing for variable detection (assuming gamma, Weibull, or lognormal distributed times to detection) produce less biased estimates of the detection probability and nominal credible interval coverage.  
% Models ignoring detection heterogeneity across subgroups yield biased estimates of detection when such heterogeneity exists, whereas models accounting for detection heterogeneity (modeled as a mixture) return reasonable coverage rates and can outperform heterogeneity-ignorant models even when there is no heterogeneity.

{\bf Keywords:} abundance; availability; N-mixture model; point counts; removal sampling; survival analysis; Bayesian

@@

In this manuscript, we develop a model for scenarios where detection rates are not constant over time. 
We consider the first time-to-detection as is done in survival analysis, defining a continuous random variable $T$ for each animal's time to first detection with a probability density function (pdf) $f_T(t)$ and cumulative distribution function (cdf) $F_T(t)$.  
We refer to the distribution of $T$ as a time-to-detection distribution (TTDD).  
One common strategy to deal with data that do not fit a constant-detection assumption is to model increased detection probability in the initial observation period via a mixture component \citep{Farnsworth2002, Farnsworth2005, EffordDawson2009, Etterson2009, Reidy2011}, although this is not yet the standard \citep{Solymos2013, Amundson2014, Reidy2016}. 
We consider the choice of whether to include a mixture component in conjunction with TTDDs with non-constant rates.

@@

Framing a model in terms of time-to-detection leads to two practical differences vis-a-vis constant-detection models.  
First, in order to model covariate and random effects on detection, we perform mixed effects linear regression on the log of the rate parameter as in \citet{Solymos2013}, whereas most existing studies instead construct regression models on the logit of the equal-interval detection probability.  
The latter is not possible when detection rates are not constant.  
Second, because we can obtain interval-specific detection probabilities from the TTDD by partitioning its cdf, we can directly model the data according to their existing interval structure rather than subdividing the observation period into intervals of equal duration.  
Indeed, with a few simplifications, our model fits exact time-to-detection data, whereas existing constant-detection removal models only approximate exact data by subdividing the observation interval into a large number of fine equal-duration intervals \citep{Reidy2011, Amundson2014}.

@@

Section \ref{sec:data} provides a description of the interval-censored time-to-detection avian point count data under consideration.
Section \ref{sec:model} introduces an N-mixture model with a generically defined TTDD for estimating abundance from removal-sampled point-count surveys.
In addition, we discuss reasonable default priors and MCMC analysis used to estimate parameters in these models.
Section \ref{sec:sim} provides three simulation studies to assess the impact of TTDD choice on estimated detection probability. 
Section \ref{sec:ovenbirds} analyzes an Ovenbird data set under different TTDDs to determine the impact of this choice on estimated detection probability and therefore estimated abundance.

@@

Our analysis is motivated by avian point-count surveys of over 50 species in Chippewa National Forest from 2008-2013 as part of the Minnesota Forest Breeding Bird Project (MNFB) \citep{Hanowski1995}.  
However, the model we describe is appropriate for point-count surveys of any species, so long as time-to-first-detection data are available.  
For our analysis, we focused on Ovenbird counts selected from one habitat type: sawtimber red pine stands with no recent logging activity.  
Each stand had up to four sites with sufficient geographical distance between sites to reduce or eliminate overlapping territories.
The data included 65 sites and a total of 381 surveys with site specific variables including site age,  stock density, and an indicator of select-/partial-cut logging during the 1990s.

Single-visit (per year) point-count surveys were conducted by trained observers at each site once annually (weather permitting).  
Fourteen different observers conducted surveys during the study period.  
69\% of surveys in our dataset involved observers in their first year at MNFB.  
Survey durations were 10 minutes, with times to first detection censored into nine intervals: a two-minute interval followed by eight one-minute intervals.  
During each survey, the Julian date, time of day, and temperature were recorded.

A total of 947 Ovenbirds were counted with a maximum of eight birds counted in any single survey.  
The counts of Ovenbirds by interval, across all sites, were 596, 69, 62, 68, 43, 33, 35, 27, and 14.

@@

Suppose that, for each survey $s$ ($s=1,\dots,S$), $N_s$ individuals are present.  
Imagine an observer could remain at the survey location until every individual is detected, marking the time to detection $t_{sb}$ (for bird, $b=1,\dots,N_s$) for each.
Then, assuming detection times for all individuals at a survey are independent, identically distributed according to a common time-to-detection distribution (TTDD), we define $T_{sb}$ as a random variable with cumulative distribution function (cdf) $F_T(t)$ and probability density function (pdf) $f_T(t)$.
However, in reality, times to first detection are truncated due to a finite survey length of $C$, meaning that each individual has a detection probability $\pdet=F_T(C)$.  
The conditional distribution of observed detection times then has pdf $f_{T|det}(t)= f_T(t)/F_T(C)$ for $0<t<C$ and cdf $F_{T|det}(t) = \int_0^t f_{T|det}(x) dx$. 
We model the number of individuals for which $t_{sb}$ is actually observed as $n_{s}^{(obs)} \ind \mbox{Binomial}\left(N_{s}, \pdet\right)$.
%$T_{sb|det} \ind f_{T|det}(t_{sb})$ where $T_{sb|det}$ is observed only for those $n_s^{(obs)}$ individuals whose observed time to first detection is less than $C$, i.e. $t_{sb}<C$. 
The instantaneous detection rate, or hazard function, is $h(t) = f_T(t) / [1-F_T(t)]$.

@@

Due to the harried process of avian point counts, times to first detection are typically not recorded exactly, but are instead censored into $I$ intervals. 
Let $C_i$ for $i=1,\dots,I$ indicate the right endpoint of the $i$th interval then $C_I$ is the total survey duration and, letting $C_0=0$, the $i$th interval is $(C_{i-1},C_{i}]$. 
Finally, let $n_{si}$ be the number of animals counted during interval $i$ on survey $s$, $n_{s}^{(obs)} = \sum_{i=1}^I n_{si}$, and $\vn_{s}=(n_{s1},\dots,n_{sI})$.
Assuming independence amongst individuals and sites, we have $\vn_{s} \ind \Mult \left(n_{s}^{(obs)}, \vp_{s}\right)$, where $\vp_{s}=(p_{s1},\dots,p_{sI})$ and is calculated from the TTDD: $p_{si} = F_{T|det}(C_i) - F_{T|det}(C_{i-1})$.

@@

For ease of reference, the final full model is provided in equation \eqref{eq:model} where the conditioning of the TTDD cdf on $\alpha$ and $\varphi_s$ is made explicit.
\begin{align}
n_s^{(obs)} &\ind \Po(\lambda_s \pdet_s) \nonumber\\
\vn_s &\ind \Mult(n_s^{(obs)}, \vp_s); \qquad \vp_s = (p_{s1},\dots,p_{sI}) \nonumber\\
\pdet_s &= F_T(C_I|\alpha,\varphi_s)\\
p_{s1} &= \left[(1-\gamma) + \gamma\Fm(C_1|\alpha,\varphi_s)\right]/\pdet_s \nonumber\\
p_{si} &= \gamma\left[\Fm(C_i|\alpha,\varphi_s) - \Fm(C_{i-1}|\alpha,\varphi_s)\right]/\pdet_s \nonumber\\
\log(\lambda_s) &= \vX_{s}^A\vbeta^A + \vZ_{s}^A\vxi^A; \qquad \xi_j^A \ind N(0,\sigma_{A[j]}^2) \nonumber\\
\label{eq:model}\log(\varphi_{s}) &= \vX_{s}^D\vbeta^D + \vZ_{s}^D\vxi^D; \qquad \xi_j^D \ind N(0,\sigma_{D[j]}^2)
\end{align}
We adopt a Bayesian approach and therefore require a prior over the model parameters.
To ease construction of a default prior for this model, we standardize all explanatory variables and then construct priors to be diffuse within a reasonable range of values.
Normal prior mean and standard deviation (sd) for the abundance intercept was set at a median abundance of 3 birds per site and a 95\% probability of 0-14 birds present (counted and uncounted).  
Normal prior mean and sd for the detection intercept were chosen so that, based on an intercept-only non-mixture model with $\alpha=1$: (i) median prior detection probability was $p_{s}^{(det)} = 0.50$, and (ii) 95\% of the prior detection probability was within $p_{s}^{(det)} \in (0.01, 1.0)$.  
Normal priors for fixed effect parameters were centered at zero with standard deviations matching the appropriate intercept term.  
All standard deviations and $\alpha$ were given half-Cauchy priors with location 0 and scale 1 for the untruncated Cauchy, and the mixture parameter $\gamma$ was assigned a Unif(0,1) prior in mixture models.
All scalar parameters were assumed independent \emph{a priori}.

We fit the models by MCMC sampling using the Bayesian statistical software Stan, implemented via the R package \texttt{rstan} version 2.8.0 \citep{Rstan2015}.  
Stan model code and \texttt{rstan} code to run the models can be found in supplemental materials.  
We discarded half of the iterations as warmup and then thinned by 10.  
We monitored convergence of the MCMC chains using Geweke z-score diagnostics \citep{Geweke1991} and reran models if lack of convergence was indicated by a non-normal distribution of the z-scores. 
We reran models if the effective sample size for any parameter was below 1000.  
The number of iterations used depended on the model and is detailed later. 
For most models, we accepted Stan defaults for initial values; however, gamma and Weibull models sometimes failed to run unless care was taken in the specification of initial values.

@@

We conducted three simulation studies to explore the behavior of models with non-constant TTDDs.
The first study compares mixture vs non-mixture models.
The second study compares the TTDD families.
In the first two studies, we utilized intercept only models to focus attention on the TTDD. 
For the third study, we included fixed and random effects for both abundance and detection and again compared the distribution families. 
In all simulation studies, we focus on accuracy in estimation of $\pdet$ which then translates into estimation of abundance. For the simulation studies following analysis we distinguish two categories of purely continuous TTDDs: peaked and nonpeaked.  
A peaked TTDD has a mode greater than zero (or $C_1$ for lognormal) while a non-peaked TTDD has a mode of zero (or less than $C_1$).

@@

To assess the need for incorporating a mixture component to increase the probability of detection in the initial interval as discussed in Section \ref{s:interval}, we simulated 16 intercept-only datasets from each of 14 TTDDs: each combination of peaked/non-peaked, mixture/non-mixture, and exponential/gamma/Weibull/lognormal, where all exponential models are non-peaked.  
We chose the number of surveys (381) and true parameter values (supplemental table \ref{S-simvals}) to mimic values from the Ovenbird analysis (Section \ref{sec:ovenbirds}).  
In particular, we set parameters such that (i) the overall expected detection probability was 0.80, (ii) for mixture datasets, $\gamma = 0.65$ (meaning 35\% of individuals were immediately detected), (iii) in nonpeaked models, 70\% of \textit{detected} individuals were observed during the first two minutes, and (iv) in peaked models, the detection mode for `hard to detect' individuals occured at 5 minutes.

@@

We summarized the analysis be calculating the average across the simulations of the posterior median of $\pdet$ (Med. $p$), the average proportion of the posterior distribution of $\pdet$ larger than the truth (Q($p$)), and average coverage for 50\% and 90\% credible intervals. If the analyses are providing a reasonable estimate of $\pdet$, we would expect Med. $p$ to be around the true value of 0.8, Q($p$) to be around 0.5 indicating that half of the posterior is above and below the true value of 0.8, and the coverage to be close to their credibility. Table \ref{tbl:sim1} provides a summary of these quantities. 
When a mixture model is used to simulate the data (bottom half of the table), there is clearly a benefit to using a mixture model for inference.  
Using a non-mixture model for inference, the credible interval coverage is near zero for most models with the exponential model overestimating $\pdet$ (Q($p$) $\approx 1$) and the other models underestimating. 
When a non-mixture model is used to simulate the data (top half of the table), there are no clearly discernable differences between the ability of a non-mixture or mixture model to capture $\pdet$. 
For nonpeaked non-mixture data (top 4 lines), the mixture model is less biased for lognormal and Weibull scenarios (Q($p$)$\approx 0.5$), and CIs are 15-40\% narrower (results not shown).
These results support the general default use of a mixture model over a non-mixture model.

@@

The previous section adressed model mis-specification in terms of the mixture component. 
Now we turn to model misspecification of the distribution family. 
We simulated 16 intercept-only datasets from the 7 different mixture TTDD models using the same settings as in the previous section, and we fit them with mixture models from each of exponential, gamma, lognormal, and Weibull families.
%Again, we ran 60,000 iterations which showed no evidence of lack of convergence according to the Geweke diagnostic and reached over 1,000 effect samples for all parameters.

@@

The previous sections studied effects of time to detection assumptions in the context of no explanatory variables.
We now incorporate fixed and random effects for abundance and detection. 
We simulated data from each of the 7 mixture TTDDs and fit models from exponential, gamma, lognormal, and Weibull mixture models.  
We simulated data using the median posterior parameter estimates obtained in the analysis of Ovenbird data in Section \ref{sec:ovenbirds}.
Because the fitted Ovenbird models did not yield peaked distributions, we simulated peaked datasets by: (i) using the same intercepts, shape parameters, and mixing parameters as for peaked data in the previous simulations, (ii) using median covariate and random effects from the Ovenbird estimates, and (iii) scaling the detection intercept and random effect to achieve true detection probabilities $\approx 0.8$ with a detection mode at 5 minutes.
See Table \ref{S-tbl:sim3} in supplement for actual parameter values.
Due to the computation time involved in estimating models with these fixed and random effects, we simulated each TTDD only once.
To obtain reasonable convergence diagnostics and effective sample sizes, these analyses ranged from 250,000-375,000 iterations.
Because of the difficulty in integrating random effects over all sites, approximate posterior distributions for the study-wide marginal $\pdet$ were obtained by simulating data from each MCMC sample and calculating the proportion of simulated Ovenbirds that were observed.

Patterns in posterior estimates of site-specific detection probabilities, $\pdet_s$, with respect to mixture and family TTDD forms were the same as in the previous simulation studies -- the inclusion of explanatory variables did not make models more robust to violations of mixture- and constant-detection assumptions (Figures \ref{S-pdet_cater_correct} and \ref{S-pdet_cater_family}).  
Posteriors for abundance fixed and random effects were the same regardless of what TTDD was assumed (Section \ref{S-s:sim3}).  
Posteriors for the mixing parameter $\gamma$ and detection fixed and random effects were the same across gamma, lognormal, and Weibull mixture models but were narrower and location-shifted for the exponential mixture model.  
%It should be remembered that detection covariate effect estimates are conditional on the estimated mixing parameter $\gamma$.  
The above patterns can also be discerned in posterior estimates from the Ovenbird data (Figure \ref{ovenposteriors}).

@@

If the estimation of effect parameters and the roles of explanatory variables are the primary interest, then our results suggest that the exact choice of TTDD may not be important.  
Abundance effect estimates are similar regardless of the chosen TTDD.
Detection effect estimates, while conditional on the mixing parameter $\gamma$, are similar across all mixture non-exponential TTDDs.  
These finding may well not hold if the same covariate is modeled in both abundance and detection models \citep{Kery2008}.

>>>>

\documentclass[useAMS,usenatbib,referee,12pt]{article}
\usepackage[margin=1.0in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{parskip}
\usepackage[round]{natbib}
\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage{caption}
\usepackage{tabularx, booktabs}
\usepackage{float}
\usepackage{adjustbox}
\usepackage[toc,page]{appendix}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{url}
\usepackage{xr}
\externaldocument[S-]{Supplement}
\doublespacing

@@

Abundance estimates from animal point-count surveys require accurate estimates of detection probabilities.  
The standard model for estimating detection from removal-sampled point-count surveys assumes that organisms at a survey site are detected at a constant rate; however, this assumption is often not justified.  
%Detection rates can be influenced by organismal behaviors, survey methodology, and observer effort.  
%Failure to account for non-constant detection leads to biased estimates of detection and therefore of abundance.  
We consider a class of N-mixture models that allows for detection heterogeneity over time through a flexibly defined time-to-detection distribution (TTDD) and allows for fixed and random effects for both abundance and detection.
%We consider a class of N-mixture models that allow for detection heterogeniety through a mixture component, flexible two-parameter time-to-detection distributions (TTDDs), and allows for fixed and random effects for both abundance and detection.
%Whereas, under the standard approach, detection probability is modeled by dividing the observation period into equal-duration intervals, we instead model the detection {\it rate} in continuous time via a time-to-detection distribution (TTDD) embedded within a hierarchical N-mixture framework.  
Our model is thus a combination of survival time-to-event analysis with unknown-N, unknown-p abundance estimation.  
We specifically explore two-parameter families of TTDDs, e.g. gamma, that can additionally include a mixture component to model increased probability of detection in the initial observation period.
We find that modeling a TTDD by using a mixture component is necessary when data have a chance of arising from a distribution of this nature.
In addition, two-parameter distributions can outperform exponential-based models both when the truth is exponential or non-exponential.
Finally, we analyze an Overbird data set from the Chippewa National Forest using mixed effect models for both abundance and detection.
We demonstrate that the effects of explanatory variables on abundance and detection are consistent across mixture TTDDs but that flexible TTDDs result in lower estimated probabilities of detection and therefore higher estimates of abundance. 
%We find that modeling a TTDD by using a mixture component for increased probability of detection in the initial observation period is necessary when data have a chance of arising from distributions of this nature.
%In addition, the flexibility offered by two-parameter distributions, e.g. gamma, can outperform exponential based models when the truth is exponential or non-exponential.
%Finally, we analyze an Overbird data set from the Chippewa National Forest using mixed effect models for both abundance and detection and demonstrate that the effects of explanatory variables on abundance and detection are consistent across mixture TTDDs and that flexible TTDDs result in lower estimated probabilities of detection and therefore higher estimates of abundance. 
% We apply this model to Ovenbird counts from the Chippewa National Forest and to datasets simulated under different time-to-detection patterns.  
% Models assuming constant detection rates produce biased estimates of detection when true detection rates vary with time, whereas models allowing for variable detection (assuming gamma, Weibull, or lognormal distributed times to detection) produce less biased estimates of the detection probability and nominal credible interval coverage.  
% Models ignoring detection heterogeneity across subgroups yield biased estimates of detection when such heterogeneity exists, whereas models accounting for detection heterogeneity (modeled as a mixture) return reasonable coverage rates and can outperform heterogeneity-ignorant models even when there is no heterogeneity.

{\bf Keywords:} abundance; availability; N-mixture model; point counts; removal sampling; survival analysis; Markov chain Monte Carlo; Stan

@@

In this manuscript, we develop a model for scenarios where detection rates are not constant over time. 
We consider the first time-to-detection as is done in survival analysis, defining a continuous random variable $T$ for each individual's time to first detection with a probability density function (pdf) $f_T(t)$ and cumulative distribution function (cdf) $F_T(t)$.  
We refer to the distribution of $T$ as a time-to-detection distribution (TTDD).  
One common strategy to deal with data that do not fit a constant-detection assumption is to model increased detection probability in the initial observation period via a mixture component \citep{Farnsworth2002, Farnsworth2005, EffordDawson2009, Etterson2009, Reidy2011}, although this is not yet the standard \citep{Solymos2013, Amundson2014, Reidy2016}. 
We consider the choice of whether to include a mixture component in conjunction with TTDDs with non-constant rates.

@@

Framing a model in terms of time-to-detection leads to two practical differences vis-a-vis constant-detection models.  
First, in order to model covariate and random effects on detection, we perform mixed effects linear regression on the log of the rate parameter as in \citet{Solymos2013}, whereas most existing studies instead construct regression models on the logit of the equal-interval detection probability.  
The latter is not possible when detection rates are not constant.  
Second, because we can obtain interval-specific detection probabilities from the TTDD by partitioning its cdf, we can directly model the data according to their existing interval structure rather than subdividing the observation period into intervals of equal duration.  
Indeed our model fits exact time-to-detection data, whereas existing constant-detection removal models only approximate exact data by subdividing the observation interval into a large number of fine equal-duration intervals \citep{Reidy2011, Amundson2014}.

@@

Section \ref{sec:data} provides a description of the interval-censored time-to-detection avian point count data under consideration.
Section \ref{sec:model} introduces an N-mixture model with a generically defined TTDD for estimating abundance from removal-sampled point-count surveys.
%In addition, we discuss reasonable default priors and MCMC analysis used to estimate parameters in these models.
Section \ref{sec:sim} provides three simulation studies to assess the impact of TTDD choice on estimated detection probability. 
Section \ref{sec:ovenbirds} analyzes an Ovenbird data set under different TTDDs to determine the impact of this choice on estimated detection probability and therefore estimated abundance.

@@

Our analysis is motivated by avian point-count surveys in Chippewa National Forest from 2008-2013 as part of the Minnesota Forest Breeding Bird Project (MNFB) \citep{Hanowski1995}.  
%However, the model we describe is appropriate for point-count surveys of any species, so long as time-to-first-detection data are available.  
For our analysis, we focused on Ovenbird counts selected from one habitat type: sawtimber red pine stands with no recent logging activity.  
Each stand had up to four sites with sufficient geographical distance between sites to reduce or eliminate overlapping territories.
The data included 65 sites and a total of 381 surveys with site specific variables including site age,  stock density, and an indicator of select-/partial-cut logging during the 1990s.

Single-visit (per year) point-count surveys were conducted by trained observers at each site once annually (weather permitting).  
Fourteen different observers conducted surveys during the study period and 69\% of surveys in our dataset involved observers in their first year at the MNFB.  
Survey durations were 10 minutes, with times to first detection censored into nine intervals: a two-minute interval followed by eight one-minute intervals.  
During each survey, the Julian date, time of day, and temperature were recorded.

% A total of 947 Ovenbirds were counted with a maximum of eight birds counted in any single survey.  
% The counts of Ovenbirds by interval, across all sites, were 596, 69, 62, 68, 43, 33, 35, 27, and 14.

While we focus on the estimation of detection probability in avian populations, the approach we describe is appropriate for point-count surveys of any species.
The methodology allows the analysis of data with 1) recorded first (possible censored) detection of each individual, 2) site-specific explanatory variables, and 3) survey-specific explanatory variables.

@@

Suppose that, for each survey $s$ ($s=1,\dots,S$), $N_s$ individuals are present.  
Imagine an observer could remain at the survey location until every individual is detected, recording the time to detection $t_{sb}$ (for bird, $b=1,\dots,N_s$) for each.
Assuming detection times for all individuals at a survey are independent, identically distributed according to a common time-to-detection distribution (TTDD), we define $T_{sb}$ as a random variable with cumulative distribution function (cdf) $F_T(t)$ and probability density function (pdf) $f_T(t)$.
In reality, times to first detection are often truncated due to a finite survey length of $C$, meaning that each individual has a detection probability $\pdet=F_T(C)$.  
The conditional distribution of observed detection times then has pdf $f_{T|det}(t)= f_T(t)/F_T(C)$ for $0<t<C$, cdf $F_{T|det}(t) = \int_0^t f_{T|det}(x) dx$, and instantaneous detection rate, or hazard function, is $h(t) = f_T(t) / [1-F_T(t)]$. 
We model the number of individuals at survey $s$ for which $t_{sb}$ is actually observed as $n_{s}^{(obs)} \ind \mbox{Binomial}\left(N_{s}, \pdet\right)$. 
%$T_{sb|det} \ind f_{T|det}(t_{sb})$ where $T_{sb|det}$ is observed only for those $n_s^{(obs)}$ individuals whose observed time to first detection is less than $C$, i.e. $t_{sb}<C$.

@@

Due to the harried process of avian point counts, times to first detection are typically not recorded exactly, but are instead censored into $I$ intervals. 
Let $C_i$ for $i=1,\dots,I$ indicate the right endpoint of the $i$th interval then $C_I$ is the total survey duration and, letting $C_0=0$, the $i$th interval is $(C_{i-1},C_{i}]$. 
Let $n_{si}$ be the number of individuals counted during interval $i$ on survey $s$, $n_{s}^{(obs)} = \sum_{i=1}^I n_{si}$, and $\vn_{s}=(n_{s1},\dots,n_{sI})$.
Assuming independence amongst individuals and sites, we have $\vn_{s} \ind \Mult \left(n_{s}^{(obs)}, \vp_{s}\right)$, where $\vp_{s}=(p_{s1},\dots,p_{sI})$ and is calculated from the TTDD: $p_{si} = F_{T|det}(C_i) - F_{T|det}(C_{i-1})$.

@@

For ease of reference, the final full model is provided in equation \eqref{eq:model} where the conditioning of the TTDD cdf on $\alpha$ and $\varphi_s$ is made explicit.
\begin{align}
n_s^{(obs)} &\ind \Po(\lambda_s \pdet_s) \nonumber\\
\vn_s &\ind \Mult(n_s^{(obs)}, \vp_s); \qquad \vp_s = (p_{s1},\dots,p_{sI}) \nonumber\\
\pdet_s &= F_T(C_I|\alpha,\varphi_s) \nonumber \\
p_{s1} &= \left[(1-\gamma) + \gamma\Fm(C_1|\alpha,\varphi_s)\right]/\pdet_s \label{eq:model} \\
p_{si} &= \gamma\left[\Fm(C_i|\alpha,\varphi_s) - \Fm(C_{i-1}|\alpha,\varphi_s)\right]/\pdet_s \nonumber\\
\log(\lambda_s) &= \vX_{s}^A\vbeta^A + \vZ_{s}^A\vxi^A; \qquad \xi_j^A \ind N(0,\sigma_{A[j]}^2) \nonumber\\
\log(\varphi_{s}) &= \vX_{s}^D\vbeta^D + \vZ_{s}^D\vxi^D; \qquad \xi_j^D \ind N(0,\sigma_{D[j]}^2) \nonumber
\end{align}
We adopt a Bayesian approach and therefore require a prior over the model parameters.
To ease construction of a default prior for this model, we standardize all explanatory variables and then construct priors to be diffuse within a reasonable range of values.
Normal prior mean and standard deviation (sd) for the abundance intercept was set at a median abundance of 3 birds per site and a 95\% probability of 0-14 birds present (counted and uncounted).  
Normal prior mean and sd for the detection intercept were chosen so that, based on an intercept-only non-mixture model with $\alpha=1$: (i) median prior detection probability was $p_{s}^{(det)} = 0.50$, and (ii) 95\% of the prior detection probability was within $p_{s}^{(det)} \in (0.01, 1.0)$.  
Normal priors for fixed effect parameters were centered at zero with standard deviations matching the appropriate intercept term.  
All standard deviations and $\alpha$ were given half-Cauchy priors with location 0 and scale 1 for the untruncated Cauchy, and the mixture parameter $\gamma$ was assigned a Unif(0,1) prior in mixture models.
All scalar parameters were assumed independent \emph{a priori}.

We fit the models by MCMC sampling using the Bayesian statistical software Stan, implemented via the R package \texttt{rstan} version 2.8.0 \citep{Rstan2015}.  
Stan model code and \texttt{rstan} code to run the models can be found in the Supplementary Materials.  
We discarded half of the iterations as warmup and then thinned by 10.  
We monitored convergence of the MCMC chains using Geweke z-score diagnostics \citep{Geweke1991} and reran models if lack of convergence was indicated by a non-normal distribution of the z-scores or if the effective sample size for any parameter was below 1000.  
The number of iterations used depended on the model and is detailed later. 
For most models, we accepted Stan defaults for initial values; however, gamma and Weibull models sometimes failed to run unless care was taken in the specification of initial values.

@@

We conducted three simulation studies to explore the behavior of models with non-constant TTDDs.
The first study compares mixture vs non-mixture models.
The second study compares the TTDD families.
In the first two studies, we utilized intercept only models to focus attention on the TTDD. 
For the third study, we included fixed and random effects for both abundance and detection and again compared the distribution families. 
In all simulation studies, we focus on accuracy in estimation of $\pdet$ which then translates into estimation of abundance. 
For the following analyses we distinguish two categories of purely continuous TTDDs: peaked and nonpeaked.  
A peaked TTDD has a mode greater than zero (or $C_1$ for lognormal) while a non-peaked TTDD has a mode of zero (or less than $C_1$).

@@

To assess the need for incorporating a mixture component to increase the probability of detection in the initial interval as discussed in Section \ref{s:interval}, we simulated 16 intercept-only datasets from each of 14 TTDDs: each combination of peaked/non-peaked, mixture/non-mixture, and exponential/gamma/Weibull/lognormal, where all exponential models are non-peaked.  
We chose the number of surveys (381) and true parameter values (Table \ref{S-simvals}) to mimic values from the Ovenbird analysis (Section \ref{sec:ovenbirds}).  
In particular, we set parameters such that (i) the overall expected detection probability was 0.80, (ii) for mixture datasets, $\gamma = 0.65$ (meaning 35\% of individuals were immediately detected), (iii) in nonpeaked models, 70\% of \textit{detected} individuals were observed during the first two minutes, and (iv) in peaked models, the detection mode for `hard to detect' individuals occured at 5 minutes.

@@

We summarized the analysis by calculating the average, across the simulations, of the posterior median of $\pdet$ (Med. $p$), the average proportion of the posterior distribution of $\pdet$ larger than the truth (Q($p$)), and average coverage for 50\% and 90\% credible intervals. If the analyses are providing a reasonable estimate of $\pdet$, we would expect Med. $p$ to be around the true value of 0.8, Q($p$) to be around 0.5 indicating that half of the posterior is above and below the true value of 0.8, and the coverage to be close to their credibility. Table \ref{tbl:sim1} provides a summary of these quantities. 
When a mixture model is used to simulate the data (bottom half of the table), there is clearly a benefit to using a mixture model for inference.  
Using a non-mixture model for inference, the credible interval coverage is near zero for most models with the exponential model overestimating $\pdet$ (Q($p$) $\approx 1$) and the other models underestimating. 
When a non-mixture model is used to simulate the data (top half of the table), there are no clearly discernable differences between the ability of a non-mixture or mixture model to capture $\pdet$. 
For nonpeaked non-mixture data (top 4 lines), the mixture model is less biased for lognormal and Weibull scenarios (Q($p$)$\approx 0.5$), and CIs are 15-40\% narrower (results not shown).
These results support the general default use of a mixture model over a non-mixture model.

@@

The previous section addressed model mis-specification in terms of the mixture component. 
Now we turn to model misspecification of the distribution family. 
We simulated 16 intercept-only datasets from the 7 different mixture TTDD models using the same settings as in the previous section, and we fit them with mixture models from each of exponential, gamma, lognormal, and Weibull families.
%Again, we ran 60,000 iterations which showed no evidence of lack of convergence according to the Geweke diagnostic and reached over 1,000 effect samples for all parameters.

@@

The previous sections studied effects of time to detection assumptions in the context of no explanatory variables.
We now incorporate fixed and random effects for abundance and detection. 
We simulated data from each of the 7 mixture TTDDs and fit models from exponential, gamma, lognormal, and Weibull mixture models.  
We simulated data using the median posterior parameter estimates obtained in the analysis of Ovenbird data in Section \ref{sec:ovenbirds}.
Because the fitted Ovenbird models did not yield peaked distributions, we simulated peaked datasets by: (i) using the same intercepts, shape parameters, and mixing parameters as for peaked data in the previous simulations, (ii) using median covariate and random effects from the Ovenbird estimates, and (iii) scaling the detection intercept and random effect to achieve true detection probabilities $\approx 0.8$ with a detection mode at 5 minutes, see Table \ref{S-tbl:sim3} for actual parameter values.
Due to the computation time involved in estimating models with these fixed and random effects, we simulated each TTDD only once.
To obtain reasonable convergence diagnostics and effective sample sizes, these analyses ranged from 250,000-375,000 iterations.
Because of the difficulty in integrating random effects over all sites, approximate posterior distributions for the study-wide marginal $\pdet$ were obtained by simulating data from each MCMC sample and calculating the proportion of simulated Ovenbirds that were observed.

The results from this simulation are qualitatively similar to that Ovenbird analysis (Figure \ref{ovenposteriors}) and thus we only briefly review the results here and provide the corresponding figures and tables in the Supplementary Material. 
Patterns in posterior estimates of site-specific detection probabilities, $\pdet_s$, with respect to mixture and family TTDD forms were the same as in the previous simulation studies -- the inclusion of explanatory variables did not make models more robust to violations of mixture- and constant-detection assumptions (Figures \ref{S-pdet_cater_correct} and \ref{S-pdet_cater_family}).  
Posteriors for abundance fixed and random effects were the same regardless of what TTDD was assumed (Section \ref{S-s:sim3}).  
Posteriors for the mixing parameter $\gamma$ and detection fixed and random effects were the same across gamma, lognormal, and Weibull mixture models but were narrower and location-shifted for the exponential mixture model.  
%It should be remembered that detection covariate effect estimates are conditional on the estimated mixing parameter $\gamma$.

@@

If the estimation of effect parameters and the roles of explanatory variables are the primary interest, then our results suggest that the exact choice of TTDD may not be important.  
Abundance effect estimates are similar regardless of the chosen TTDD.
Detection effect estimates, while conditional on the mixing parameter $\gamma$, are similar across all mixture non-exponential TTDDs.  
These findings may well not hold if the same covariate is modeled in both abundance and detection models \citep{Kery2008}.
